Using Genetic Algorithms to Find Approximations for the Minimum Vertex Cover Problem

### 2018-02-28 paper.tex
done for today
done for almost everything

% including packages
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{hyperref}
\usepackage[all]{hypcap}
\usepackage[autostyle]{csquotes}
\usepackage{biblatex}
\usepackage{amsfonts}
\usepackage{algorithm}
\usepackage[end]{algpseudocode}
\usepackage{caption}
\usepackage{setspace}

@@

% cover page
\pagenumbering{roman}
\cleardoublepage\phantomsection\addcontentsline{toc}{section}{Cover Page}

@@

% abstract page
\cleardoublepage\phantomsection\addcontentsline{toc}{section}{Abstract}

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod
tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam,
quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo
consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse
cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}

@@

% table of contents page
\cleardoublepage\phantomsection\addcontentsline{toc}{section}{Contents}
\tableofcontents
\pagebreak

% List of Figures and Tables
\cleardoublepage\phantomsection\addcontentsline{toc}{subsection}{List of Figures}
\listoffigures

@@

\section{Introduction}
Once the NP-hardness of a combinatorial optimization
problem is established, the search for an optimal solution
is abandoned. The goal then becomes one of finding a
good heuristic, i.e. a polynomial running time algorithm
that can find solutions close to the optimal. In most cases,
traditional heuristics are problem dependent; a heuristic is
tailored to the specific problem it is trying to solve.

In this paper, we present an alternative approach that
uses genetic algorithms as a generalized heuristic for solving NP-hard combinatorial optimization problems. The
application of a genetic algorithm is demonstrated here for
the \textit{minimum vertex cover} problem. These algorithms
have been successfully applied to a broad range of problems. This wide range can be tackled by genetic algorithms
mainly due to the fact that they work with an encoding of
the domain rather than with the problem domain itself.

\section{Genetic Algorithms}
\subsection{The Intuition Behind GAs}
Genetic Algorithms (GAs) \cite{coley, goldberg} are population based
search algorithms, where by repeated use of genetic operations,
such as \textem{mutation}, \textem{selection}, \textem{crossover}, etc...
Successive new generations of better populations in the direction of
search objectives are created.
They are inspired by Darwinian principles
based on natural evolution.
In other words, the main idea behind genetic
algorithms is that only the \textem{fittest} individuals will survive
among generations of evolution.

@@

\item
a \textem{mutation} operator to avoid permanent loss of diversity within the
solutions.
This allows introducing new information in the population.

@@

\begin{algorithm}
\caption{GenticAlgorithm$\left( n, u, p_c, p_m, f \right)$}
\label{alg:ga}

@@

\section{The Minimum Vertex Cover Problem}
The problem of finding a minimum vertex cover is a classical
optimization problem in computer science and is a typical
example of an NP-complete optimization problem \cite{wolfram:mvc}
that has approximation algorithms.
The mvcp of an undirected graph $G = (V, E)$ where $V$ is
the set of vertices and $E$ denotes the set of edges,
consists in finding the smallest subset $V' \subseteq V$ such that
$\forall \langle i, j \rangle \in E$, we have $i \in V'$ or $j \in V'$ (or both).
$V'$ is said to be a vertex cover of $G$.

@@

An interesting property of the mvcp is that by
solving it, we also have a solution for two other
graph problems: the \textit{maximum independent set problem}
(given $G = (V, E)$, find the largest subset $V' \subseteq V$
such that $\forall i, j \in V', \langle i, j \rangle \notin E$)
and the \textit{maximum clique problem}
(given $G = (V, E)$, find the largest subset $V' \subseteq V$
such that $\forall i, j \in V', \langle i, j \rangle \in E$)
The close relationship between these problems is characterized
by the following lemma \cite{6:misp}.
\begin{lemma}
\ \\[3mm]
For any graph $G = (V, E)$ and $V' \subseteq V$, the following
statements are equivalent:
\begin{itemize}
\vspace{-3mm}
\setlength{\parskip}{0.5em}
\item $V'$ is the minimum vertex cover of $G$.
\item $V - V'$ is the maximum independent set in $G$.
\item $V - V'$ is the maximum clique in $\bar{G} = (V, \bar{E})$,\\[1mm]
where
$\bar{E}=\left\{ \langle i, j \rangle\in V \times V :
\langle i, j \rangle \notin E \right\}$
\end{itemize}
\end{lemma}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.5\textwidth]{ps}
\caption[The regular graph after Papadimitriou
and Steiglitz]{Construction of the regular graph after Papadimitriou
and Steiglitz with $k=3$.}
\label{fig:ps}
\end{figure}

@@

\begin{algorithm}
\caption{vercov($G = (V, E)$)}
\label{alg:vercov}
\begin{algorithmic}[1]

@@

\section{Experiments}
For the experimental tests,
the genetic algorithm software package GENEsYs is used \cite{1:misp}.
This implementation is based on the widely used GENESIS software by Grefenstette \cite{5:mvcp},
but allows for more  flexibility concerning genetic operators and data monitoring.
The parameter settings for the experiments are
given in section~\ref{sec:GAmvcp}.

@@

\begin{algorithm}
\caption{GenerateRandomGraph$(n, d, k)$}
\label{alg:gen}
\begin{algorithmic}[1]
\algnotext{EndFor}
\algnotext{EndIf}

@@

\paragraph{Table~\ref{tbl:t2}:}
For each of the problems in the test set~1, a total of $N=100$
independent runs of the vercov algorithm is performed.
The results are summarized in table~\ref{tbl:t2}.
Notice that for one problem instance, the results may differ in each run due to the
non-determinism of the algorithm (see algorithm~\ref{alg:vercov} line~\ref{alg:rand}).

\begin{table}[!htbp]
\centering
\includegraphics[width=1\textwidth]{t2}
\caption[Results obtained by vercov algorithm for graphs in test set~1]{%
Experimental results obtained by the vercov algorithm for five random graphs of size $n = 100$ with edge density: $d = 0.1$ (``mvcp100-01''), $d = 0.2$ (``mvcp100-02''), $d = 0.3$ (``mvcp100-03''), $d = 0.5$ (``mvcp100-04''), $d = 0.5$ (``mvcp100-05'') and the regular graph of size $n = 100 \ (k=32)$ from Papadimitriou and Steiglitz (``PS100'').%
}
\label{tbl:t2}
\end{table}

\paragraph{Test set~2:}
To test the behaviour of the genetic algorithm
as well as the vercov algorithm for an even
larger problem size, The same experiments
were performed for larger graphs.
The results in table~\ref{tbl:t3}
and table~\ref{tbl:t4} correspond to test set~2,
which consists of five random graphs of size
$n = 200$ with different edge densities $d$.
Densities $d \in \{0.1, 0.2, 0.3, 0.4, 0.5\}$ are used.
Moreover, we construct the graphs in a way that guarantees an optimum of quality 110
(see algorithm~\ref{alg:gen}, we set $k = 110$).
In addition to a graph instance of the regular graphs introduced by
Papadimitriou and Steiglitzof of size $n = 220 \ (k=66)$.
In this case, the genetic algorithm
was allowed to run for $4 \cdot 10^4$ function evaluations.\\
The corresponding results obtained by the
genetic algorithm are shown in table~\ref{tbl:t3}, while the results
from the vercov algorithm are shown in table~\ref{tbl:t4}.
See the previous paragraphs for a better understanding of these results.

@@

\begin{table}[!htbp]
\centering
\includegraphics[width=1\textwidth]{t4}
\caption[Results obtained by vercov algorithm for graphs in test set~2]{%
Experimental results obtained by the vercov algorithm for five random graphs of size $n = 200$ with edge density: $d = 0.1$ (``mvcp200-01''), $d = 0.2$ (``mvcp200-02''), $d = 0.3$ (``mvcp200-03''), $d = 0.5$ (``mvcp200-04''), $d = 0.5$ (``mvcp200-05'') and the regular graph of size $n = 202 \ (k=66)$ from Papadimitriou and Steiglitz (``PS202'').%
}
\label{tbl:t4}
\end{table}

\subsection{Experimental Conclusions}
From these tables, a number of interesting conclusions
can be drawn.
For the random graphs used to compare the
genetic algorithm and the vercov algorithm,
it is known by construction that an optimum of
quality 55 exists for test set~1,
(respectively, an optimum of quality 110 exists
for the test set~2).
This optimum is likely to be the global optimum.
And it is found by the genetic algorithm at least once
within the $N = 100$ runs that were performed.

Moreover, the randomly constructed problems quickly
become simpler for the genetic algorithm when the edge
density is increased.
For an edge density of $d \in \{0.4, 0.5\}$, the
genetic algorithm almost surely finds the global optimum
of the problems, while vercov algorithm gives
a relatively terrible results.

In case of the regular graph after Papadimitriou and
Steiglitz, the genetic algorithm is able to find the global
optimum in about $2/3$ of all runs, while the remaining
runs identify a solution of quality $2k + 2$.
Whereas, vercov algorithm gives a solution of
quality $2k + 2$ in all of the 100 runs as expected \cite{10:mvcp}.
Recall that the graphs are constructed
so as to force the vercov algorithm into yielding solutions
of quality $2k + 2$.
The random choice of edges (line~\ref{alg:rand} of algorithm~\ref{alg:vercov})
implies that nodes from either layers one and two, or layers two and three
are involved in the solution being constructed,
which in turn implies that two layers have to be part of the solution
found by the algorithm.

For the randomly constructed graphs, the vercov algorithm
performs poorly.
It barely finds solutions of quality 80 (for graphs of size 100)
and 172 (for graphs of size 200).
The average performance of the vercov algorithm (surprisingly)
decreases as the edge density is increased;
i.e. the problems become even harder for the vercov algorithm
while they become much simpler for the genetic algorithm when
the edge density increases.
The average fitness found by the genetic algorithm is notably
better than the one found by the vercov algorithm.
One can demonstrate this by calculating
the ratio of the absolute (or equivalently, the relative) error.
For example, for the first problem instance,
let $\bar{f}_1, \bar{f}_2$ be the average found by
the genetic algorithm, vercov algorithm respectively.
We have $ (\bar{f}_2 - 55) / (\bar{f}_1 - 55) = (86.46 - 55)/(62.61 - 55) \approx 4 = 400\%$
that is, the error was reduced by a factor of 4.

\section{Conclusion}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod
tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam,
quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo
consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse
cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\cite[hi][go]{coley} \cite{davis}

@@

\cleardoublepage\phantomsection\addcontentsline{toc}{section}{References}
%\bibliographystyle{abbrv}
%\bibliography{references}
\printbibliography

>>>>

% including packages
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage[autostyle]{csquotes}
\usepackage{biblatex}
\usepackage{amsfonts}
\usepackage[end]{algpseudocode}
\usepackage{caption}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{algorithm}
\usepackage[all]{hypcap}

@@

% cover page
\pagenumbering{roman}
\phantomsection\addcontentsline{toc}{section}{Cover Page}

@@

% abstract page
\phantomsection\addcontentsline{toc}{section}{Abstract}

\begin{abstract}
Remarkable results obtained from the application of a
genetic algorithm to the minimum vertex cover problem
are reported, and tested against the results of the
best known heuristic algorithm for graphs of sizes 100 and 200 nodes.
In contrast to many other approximation algorithms that use problem-specific
knowledge, the approach presented relies on a graded penalty term component
of the fitness function to penalize infeasible solutions.
\end{abstract}

@@

% table of contents page
\setstretch{0.5}
\phantomsection\addcontentsline{toc}{section}{Contents}
\tableofcontents
\setstretch{1.0}

% List of Figures and Tables
\phantomsection\addcontentsline{toc}{subsection}{List of Figures}
\listoffigures

@@

\section{Introduction}
\label{sec:intro}
Once the NP-hardness of a combinatorial optimization
problem is established, the search for an optimal solution
is abandoned.
Our goal then becomes one of finding good heuristics
for that problem.
That is, we try to find approximation algorithms
that have polynomial time complexities and return near-optimal
solutions.
In most cases, the heuristic is problem dependent; it is tailored to specific problem
it is trying to solve, and it is tightly coupled with observations made about the problem.
The \textit{minimum vertex cover} problem is a typical example of such
problems \cite{wolfram:mvc}.
Due to its numerous applications especially in various
matching problems, the problem is not abandoned and it has many
approximation algorithms.

In this work, we present an alternative approach that
uses \textit{genetic algorithms} as a generalized heuristic for
solving NP-hard combinatorial optimization problems.
These algorithms have been successfully applied to a broad
range of problems.
This wide range can be tackled by genetic algorithms
mainly due to the fact that these algorithms work with an
encoding of the problem rather than the characteristics
of that problem.
Remarkable results were obtained and compared to the results of the
best known heuristic algorithm for the mvcp.

The outline of this paper is as follows: Firstly, Section~\ref{sec:ga} gives a
short introduction to genetic algorithms, and explains briefly how they work.
Section~\ref{sec:mvcp} introduces the mvcp with two of its main heuristics,
and explains its representation for an application of the genetic algorithm.
Lastly, Section~\ref{sec:exper} presents our experimental results and
points out the notable observations that can be drawn from these results.

\section{Genetic Algorithms}
\label{sec:ga}
\subsection{The Intuition Behind GAs}
Genetic Algorithms (GAs) \cite{coley, goldberg} are population based
search algorithms, where by repeated use of genetic operations,
such as \textem{mutation}, \textem{selection}, \textem{crossover}, etc...
Successive new generations of better populations in the direction of
search objectives are created.
They are inspired by Darwinian principles
based on natural evolution.
In other words, the main idea behind genetic
algorithms is that only the \textem{fittest} individuals will survive
among generations of evolution.

@@

@@

\begin{algorithm}[!htbp]
\caption{GenticAlgorithm$\left( n, u, p_c, p_m, f \right)$}
\label{alg:ga}

@@

\section{The Minimum Vertex Cover Problem}
\label{sec:mvcp}
The problem of finding a minimum vertex cover is a classical
optimization problem in computer science and is a typical
example of an NP-complete optimization problem \cite{wolfram:mvc}
that has approximation algorithms.
The mvcp of an undirected graph $G = (V, E)$ where $V$ is
the set of vertices and $E$ denotes the set of edges,
consists in finding the smallest subset $V' \subseteq V$ such that
$\forall \langle i, j \rangle \in E$, we have $i \in V'$ or $j \in V'$ (or both).
$V'$ is said to be a vertex cover of $G$.

@@

An interesting property of the mvcp is that by
solving it, we also have a solution for two other
graph problems: the \textit{maximum independent set problem}
(given $G = (V, E)$, find the largest subset $V' \subseteq V$
such that $\forall i, j \in V', \langle i, j \rangle \notin E$)
and the \textit{maximum clique problem}
(given $G = (V, E)$, find the largest subset $V' \subseteq V$
such that $\forall i, j \in V', \langle i, j \rangle \in E$)
The close relationship between these problems is characterized
by the following lemma \cite{6:misp}.
\begin{lemma}
\label{lem}
\ \\[3mm]
For any graph $G = (V, E)$ and $V' \subseteq V$, the following
statements are equivalent:
\begin{itemize}
\vspace{-3mm}
\setlength{\parskip}{0.5em}
\item $V'$ is the minimum vertex cover of $G$.
\item $V - V'$ is the maximum independent set in $G$.
\item $V - V'$ is the maximum clique in $\bar{G} = (V, \bar{E})$,\\[1mm]
where
$\bar{E}=\left\{ \langle i, j \rangle\in V \times V :
\langle i, j \rangle \notin E \right\}$
\end{itemize}
\end{lemma}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.5\textwidth]{ps}
\caption[The regular graph after Papadimitriou
and Steiglitz for $k=3$]
{Construction of the regular graph after Papadimitriou
and Steiglitz with $k=3$}
\label{fig:ps}
\end{figure}

@@

\begin{algorithm}[!htbp]
\caption{vercov($G = (V, E)$)}
\label{alg:vercov}
\begin{algorithmic}[1]

@@

\section{Experiments}
\label{sec:exper}
For the experimental tests,
the genetic algorithm software package GENEsYs is used \cite{1:misp}.
This implementation is based on the widely used GENESIS software by Grefenstette \cite{5:mvcp},
but allows for more  flexibility concerning genetic operators and data monitoring.
The parameter settings for the experiments are
given in section~\ref{sec:GAmvcp}.

@@

\begin{algorithm}[!htbp]
\caption{GenerateRandomGraph$(n, d, k)$}
\label{alg:gen}
\begin{algorithmic}[1]
\algnotext{EndFor}
\algnotext{EndIf}

@@

\paragraph{Table~\ref{tbl:t2}:}
For each of the problems in the test set~1, a total of $N=100$
independent runs of the vercov heuristic is performed.
The results are summarized in table~\ref{tbl:t2}.
Notice that for one problem instance, the results may differ in each run due to the
non-determinism of the algorithm (see algorithm~\ref{alg:vercov} line~\ref{alg:rand}).

\begin{table}[!htbp]
\centering
\includegraphics[width=1\textwidth]{t2}
\caption[Results obtained by vercov heuristic for graphs in test set~1]{%
Experimental results obtained by the vercov heuristic for five random graphs of size $n = 100$ with edge density: $d = 0.1$ (``mvcp100-01''), $d = 0.2$ (``mvcp100-02''), $d = 0.3$ (``mvcp100-03''), $d = 0.5$ (``mvcp100-04''), $d = 0.5$ (``mvcp100-05'') and the regular graph of size $n = 100 \ (k=32)$ from Papadimitriou and Steiglitz (``PS100'').%
}
\label{tbl:t2}
\end{table}

\paragraph{Test set~2:}
To test the behaviour of the genetic algorithm
as well as the vercov heuristic for an even
larger problem size, The same experiments
were performed for larger graphs.
The results in table~\ref{tbl:t3}
and table~\ref{tbl:t4} correspond to test set~2,
which consists of five random graphs of size
$n = 200$ with different edge densities $d$.
Densities $d \in \{0.1, 0.2, 0.3, 0.4, 0.5\}$ are used.
Moreover, we construct the graphs in a way that guarantees an optimum of quality 110
(see algorithm~\ref{alg:gen}, we set $k = 110$).
In addition to a graph instance of the regular graphs introduced by
Papadimitriou and Steiglitzof of size $n = 220 \ (k=66)$.
In this case, the genetic algorithm
was allowed to run for $4 \cdot 10^4$ function evaluations.\\
The corresponding results obtained by the
genetic algorithm are shown in table~\ref{tbl:t3}, while the results
from the vercov heuristic are shown in table~\ref{tbl:t4}.
See the previous paragraphs for a better understanding of these results.

@@

\begin{table}[!htbp]
\centering
\includegraphics[width=1\textwidth]{t4}
\caption[Results obtained by vercov heuristic for graphs in test set~2]{%
Experimental results obtained by the vercov heuristic for five random graphs of size $n = 200$ with edge density: $d = 0.1$ (``mvcp200-01''), $d = 0.2$ (``mvcp200-02''), $d = 0.3$ (``mvcp200-03''), $d = 0.5$ (``mvcp200-04''), $d = 0.5$ (``mvcp200-05'') and the regular graph of size $n = 202 \ (k=66)$ from Papadimitriou and Steiglitz (``PS202'').%
}
\label{tbl:t4}
\end{table}

\subsection{Experimental Conclusions}
From these tables, a number of interesting conclusions
can be drawn.
For the random graphs used to compare the
genetic algorithm and the vercov heuristic,
it is known by construction that an optimum of
quality 55 exists for test set~1,
(respectively, an optimum of quality 110 exists
for the test set~2).
This optimum is likely to be the global optimum.
And it is found by the genetic algorithm at least once
within the $N = 100$ runs that were performed.

Moreover, the randomly constructed problems quickly
become simpler for the genetic algorithm when the edge
density is increased.
For an edge density of $d \in \{0.4, 0.5\}$, the
genetic algorithm almost surely finds the global optimum
of the problems, while vercov heuristic gives
a relatively terrible results.

In case of the regular graph after Papadimitriou and
Steiglitz, the genetic algorithm is able to find the global
optimum in about $2/3$ of all runs, while the remaining
runs identify a solution of quality $2k + 2$.
Whereas, vercov heuristic gives a solution of
quality $2k + 2$ in all of the 100 runs as expected \cite{10:mvcp}.
Recall that the graphs are constructed
so as to force the vercov heuristic into yielding solutions
of quality $2k + 2$.
The random choice of edges (line~\ref{alg:rand} of algorithm~\ref{alg:vercov})
implies that nodes from either layers one and two, or layers two and three
are involved in the solution being constructed,
which in turn implies that two layers have to be part of the solution
found by the algorithm.

For the randomly constructed graphs, the vercov heuristic
performs poorly.
It barely finds solutions of quality 80 (for graphs of size 100)
and 172 (for graphs of size 200).
Moreover, notice that the quality of its solutions are dominated
by the quality of the solutions found by the genetic algorithm;
the worst solutions found by the genetic algorithm are almost as
good as the best solutions found by the vercov heuristic.

The average performance of the vercov heuristic (surprisingly)
decreases as the edge density is increased;
i.e. the problems become even harder for the vercov heuristic
while they become much simpler for the genetic algorithm when
the edge density increases.
The average fitness found by the genetic algorithm is notably
better than the one found by the vercov heuristic.
One can demonstrate this by calculating
the ratio of the absolute (or equivalently, the relative) error.
For example, for the first problem instance,
let $\bar{f}_1, \bar{f}_2$ be the average found by
the genetic algorithm, vercov heuristic respectively.
We have $ (\bar{f}_2 - 55) / (\bar{f}_1 - 55) = (86.46 - 55)/(62.61 - 55) \approx 4 = 400\%$
that is, the error was reduced by a factor of 4.

\section{Conclusion}
\label{sec:conc}
This work has demonstrated that genetic algorithms can
be used in a fairly straightforward way to find good
approximative solutions for NP-hard problems, namely
the minimum vertex cover problem.
The robustness of the presented solution along with
the positive results obtained support the idea
that genetic algorithms are a desirable approach for tackling
these type of problems.
To tackle a new problem, rather than having to construct
tailored heuristics to handle the problem under consideration,
we advocate the use of genetic algorithms where the
only change to perform is the formulation of a new fitness
function.

@@

\phantomsection\addcontentsline{toc}{section}{References}
%\bibliographystyle{abbrv}
%\bibliography{references}
\printbibliography
